Windowed least-squares model reduction for dynamical systems

### 2019-10-07 section4/spatial_numerical.tex
latest

\subsection{\spatialAcronym: direct solution approach}\label{sec:direct} 
Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.

In the present context, the most straightforward direct solution approach consists of the 
following steps:
\begin{enumerate}
\item Numerically discretize the FOM ODE (and hence the \textit{integrand} of~\eqref{eq:obj_gen_slab}).
\item Select a numerical quadrature rule to evaluate the \textit{integral}
in~\eqref{eq:obj_gen_slab}.
\end{enumerate}
To achieve this agenda, first define a discrete temporal grid 
by dividing each time window $[\timeStartArg{n},\timeEndArg{n}]$ into
$\nstepsArg{n}$ strictly increasing time-step instances described by
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}$ with  
\begin{align}
&\timeWindowArg{n}{0} \le \ldots \le \timeWindowArg{n}{\nstepsArg{n}} , \label{eq:timegrid1} \;
\timeWindowArg{n}{0} \defeq \timeStartArg{n},\;
\timeWindowArg{n}{\nstepsArg{n}} \defeq \timeEndArg{n} . 
%\timeWindowArg{n}{0} = \timeStartArg{n}, \timeWindowArg{n}{\nstepsArg{n}} =
%\timeEndArg{n}
\end{align}
This discretization is depicted in Figure~\ref{fig:slab_fig2}.
We now outline the direct solution approach for linear-multistep schemes; we note that the formulation for
other time-stepping methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
\caption{Decomposition of time windows into time-step instances.} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

\objectiveArgLMS{n} &\vcentcolon (\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}}) \mapsto
\frac{1}{2} \sum_{i=1}^{\nstepsArg{n}} \quadWeightsLMSScalarArg{n}{i} [\residArg{n,i}(\stateyDiscreteArgnt{n,i},\ldots, \stateyDiscreteArgnt{n,i-k(n)})]^T  \stweightingMatArgt{n}{\timeWindowArg{n}{i}} \residArg{n,i}(\stateyDiscreteArgnt{n,i},\ldots, \stateyDiscreteArgnt{n,i-k(n)}), \\

@@

Over each window, \methodAcronym\ computes the generalized coordinates $\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$, for $n=1,\ldots,\nslabs$, as the solution to the minimization problems,

@@

&\underset{\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}} \in \trialspace}{\text{minimize } }
\objectiveArgLMS{n} (\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}}) \\
& \text{subject to} \hspace{0.2 in}  \stateyDiscreteArg{n,0} =
\begin{cases} \approxstateDiscreteArg{n-1,\nstepsArg{n-1}} & n = 2,\ldots,\nslabs,\\
\approxstateIC & n=1. \end{cases} \end{align*}

@@

Defining $\genstateDiscreteArg{n,i} \in \RR{\romdim}$ to be the discrete approximation to the generalized coordinates at the $i$th time-step instance over the $n$th window, 

@@

& \underset{\genstateyDiscreteArg{n,0},\ldots,\genstateyDiscreteArg{n,\nstepsArg{n}}}{\text{minimize } }
\objectiveArgLMS{n} (\basisspace \genstateyDiscreteArg{n,0} + \stateIntercept,\ldots,\basisspace \genstateyDiscreteArg{n,\nstepsArg{n}} + \stateIntercept) \\ 

@@

\genstateyDiscreteArg{n,0} =
\begin{cases} \genstateDiscreteArg{n-1, \nstepsArg{n-1}} & n = 2,\ldots,\nslabs \\
\basisspace^T(\stateFOMIC - \stateIntercept)& n=1. \end{cases} 

@@

 = \underset{\genCollocMatyArg{n}{1},\ldots,\genCollocMatyArg{n}{\nstepsArg{n}} }{\text{argmin } }
\objectiveArgC{n} (\basisspace \genCollocMatyArg{n}{1} + \stateInterceptMat,\ldots,\basisspace \genCollocMatyArg{n}{\nstepsArg{n}} + \stateInterceptMat) \qquad \text{ for } n = 1,2,...,\nslabs,

@@

\basisspace^T (\approxstateIC - \stateIntercept)& n=1, \;  i=1,

@@

To describe the Gauss--Newton method, we first define a ``vectorization" function that takes a tuple of time-local solutions at each time-step instance and outputs 
a concatenated vector, 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
for arbitrary $p$ and $m$. The stacked vector of generalized coordinates over the $n$th time window are then defined as, 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,0},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
Now, define the weighted space--time residual over the entire window as
\begin{equation*}
\residLMSSlabArg{n} : \genstatecollocMatySlabArg{n} \mapsto \begin{bmatrix}
 \sqrt{\quadWeightsLMSScalarArg{n}{1} } \stweightingMatOneArg{n,1} \residLMSArg{n,1}( \basisspace \genstateyDiscreteArgnt{n,1} + \stateIntercept, \ldots,\basisspace \genstateyDiscreteArgnt{n,1 - k^n(1) }  + \stateIntercept) \\
\vdots \\
 \sqrt{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}} } \stweightingMatOneArg{n,\nstepsArg{n}} \residLMSArg{n,\nstepsArg{n}}( \basisspace \genstateyDiscreteArgnt{n,\nstepsArg{n}} + \stateIntercept, \ldots,\basisspace \genstateyDiscreteArgnt{n,\nstepsArg{n} - k^n(\nstepsArg{n}) } + \stateIntercept) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{n} \defeq \unroll (\genstateyDiscreteArg{n,0},\ldots,\genstateyDiscreteArg{n,\nstepsArg{n}} ).$
It is worth noting that, by design,
\begin{equation*}
\objectiveArgLMS{n} \bigg( \basisspace \genstateDiscreteArg{n,0} + \stateIntercept,\ldots,\basisspace \genstateDiscreteArg{n,\nstepsArg{n}} + \stateIntercept \ \bigg) 
\equiv
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard Gauss--Newton method. The algorithm consists of three fundamental steps: (1) compute the FOM O$\Delta$E residual given the current guess, (2) compute the time window residual-Jacobian and form the normal equations, and (3) solve the normal equations and update the state.

The practical implementation of the Gauss--Newton algorithm requires an efficient method for computing the time window residual-Jacobians. In particular, it is important to note that for linear multistep methods (and many other time marching schemes) the time window residual Jacobian will (almost) be a block-diagonal 
sparse matrix with the sparsity pattern,
\begin{equation*}
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{n}}(\genstatecollocMatSlabArg{n})  =
 \begin{bmatrix*}[l]
\matshapeb & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*}.
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
This sparsity pattern can be leveraged to assemble the time window residual-Jacobian from local residual Jacobians. 
Further, the sparsity pattern can be 
leveraged to efficiently compute the Jacobian matrix-matrix product in the normal equations.
It is also worth noting that the 
normal equations will also (almost) be block-diagonal; this is a fact that can be leveraged to speed up the linear solves at each Gauss--Newton iteration.

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
The principle cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations are several such examples. Further, the space--time formulation exposes an extra dimension for parallelization that can be exposed to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $tol$; initial guess, $\genstateGuessDiscreteArg{n,0}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
%\While{ $tol \le \frac{1}{2} \sum_{i=1}^{\nstepsArg{n}} \sum_{j=1}^{\ncollocArg{n}{i}} \big[ \dot{\state}
%(\collocPointArg{n}{i}{j})  - \velocity(\stateArg{}{\collocPointArg{n}{i}{j}}) \big]^T \stweightingMatArg{n}(\collocPointArg{n}{i}{j}) \big[ \dot{\state}(\collocPointArg{n}{i}{j}) - \velocity(\stateArg{}{\collocPointArg{n}{i}{j}}) \big] \int_{\timeStartArg{n}}^{\timeEndArg{n}}   \lagrangeBasisScalarArg{n}{i}{j}(t) d \tau$}
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$k \leftarrow 0$ \Comment{Set counter}\\
$\genstatecollocMatSlabArg{n}_k \leftarrow \unroll(\genstateGuessDiscreteArg{n,0}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
\While{\text{converged} == \text{false}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{n}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
$\Delta \genstatecollocMatSlabArg{n} \solveAlgo [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
\If{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \text{tol}$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
$k\leftarrow k+1$
}
\end{algorithm}

>>>>

\subsection{\spatialAcronym: direct solution approach}\label{sec:direct} 
\KTC{Does the time grid definted in this section have any relationship to the
time grid used for time discretization? Haven't gotten past HERE below, but
wanted to check (maybe it's made clear later).}
\EP{This time grid becomes the time discretization over each window. At this point in the paper, 
there is no notion of a time grid for \methodAcronym}

Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
\KTC{Passive voice:} \EP{I think this reads clearly} The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.  
\KTC{is this a
spectral method, or the spectral method? clarify} \EP{I don't understand what you mean. What is ``the" spectral method}

\KTC{you can't devide (i.e., decompose) a window into instances. you can divide a window into
sub-windows. you can also simply define a set of time instances in the window.
no need to emphasize `strictly increasing', as any set of points in the window
will by necessity be ordered.}

In the present context, the most straightforward direct solution approach consists of the 
following steps: (1) numerically discretize the FOM ODE (and hence the \textit{integrand} of~\eqref{eq:obj_gen_slab}) and 
(2) select a numerical quadrature rule to evaluate the \textit{integral} in~\eqref{eq:obj_gen_slab}.
To achieve this agenda, first define a discrete temporal grid 
by defining
$\nstepsArg{n}+1$ strictly increasing time instances described by
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}$ with
%\KTC{use cdots, not
%hdots or ldots. Don't use $\defeq$ for the first and last times; these are
%just equalities, i.e., you happen to set them equal to those values, but they
%aren't identically the same:}
\begin{align}
&\timeWindowArg{n}{0} \le \ldots \le \timeWindowArg{n}{\nstepsArg{n}} , \label{eq:timegrid1} \;
\timeWindowArg{n}{0} = \timeStartArg{n},\;
\timeWindowArg{n}{\nstepsArg{n}} = \timeEndArg{n} . 
%\timeWindowArg{n}{0} = \timeStartArg{n}, \timeWindowArg{n}{\nstepsArg{n}} =
%\timeEndArg{n}
\end{align}
Figure~\ref{fig:slab_fig2} depicts this discretization.
We now outline the direct solution approach for linear-multistep schemes; we note that the formulation for
other time-stepping methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
	\caption{Partitioning of time windows into time-step instances. \KTC{Don't call them `time-step instances'.
	This woudl be an instance of a time step. These are `time instances',
	correct? If so, fix this terminology everywhere.}} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

\objectiveArgLMS{n} &\vcentcolon (\stateyDiscreteArgnt{0},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \mapsto
\frac{1}{2} \sum_{i=1}^{\nstepsArg{n}} \quadWeightsLMSScalarArg{n}{i} [\residArg{n,i}(\stateyDiscreteArgnt{i},\ldots, \stateyDiscreteArgnt{i-k^n(i)})]^T  \stweightingMatArgt{n}{\timeWindowArg{n}{i}} \residArg{n,i}(\stateyDiscreteArgnt{i},\ldots, \stateyDiscreteArgnt{i-k^n(i)}), \\

@@

Over each window, \directMethodAcronym\ computes the generalized coordinates $\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$, for $n=1,\ldots,\nslabs$, as the solution to the minimization problems,

@@

&\underset{\stateyDiscreteArgnt{0},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}} \in \trialspace}{\text{minimize } }
\objectiveArgLMS{n} (\stateyDiscreteArgnt{0},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \\
& \text{subject to} \hspace{0.2 in}  \stateyDiscreteArg{0} =
\begin{cases} \spatialIC & n = 2,\ldots,\nslabs,\\
\genstateICOne & n=1. \end{cases} \end{align*}

@@

Defining $\genstateDiscreteArg{n,i} \in \RR{\romdim}$ to be the discrete approximation to the generalized coordinates at the $i$th time instance over the $n$th window, 

@@

& \underset{\genstateyDiscreteArg{0},\ldots,\genstateyDiscreteArg{\nstepsArg{n}}}{\text{minimize } }
\objectiveArgLMS{n} (\basisspaceArg{n} \genstateyDiscreteArg{0} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateyDiscreteArg{\nstepsArg{n}} + \stateInterceptArg{n}) \\ 

@@

\genstateyDiscreteArg{0} =
\begin{cases} \spatialIC & n = 2,\ldots,\nslabs \\
\genstateICOne& n=1. \end{cases} 

@@

 = \underset{\genCollocMatyArg{}{1},\ldots,\genCollocMatyArg{}{\nstepsArg{n}} }{\text{argmin } }
\objectiveArgC{n} (\basisspaceArg{n} \genCollocMatyArg{}{1} + \stateInterceptMat,\ldots,\basisspaceArg{n} \genCollocMatyArg{}{\nstepsArg{n}} + \stateInterceptMat) \qquad \text{ for } n = 1,2,...,\nslabs,

@@

\basisspaceTArg{n} (\approxstateIC - \stateIntercept)& n=1, \;  i=1,

@@

To describe the Gauss--Newton method, we first define a ``vectorization" function that takes a tuple of time-local solutions at each time-step instance and outputs 
a concatenated vector, 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
for arbitrary $p$ and $m$. The stacked vector of generalized coordinates over the $n$th time window are then defined as, 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,1},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
Now, define the weighted space--time residual over the entire window as
\begin{equation*}
\residLMSSlabArg{n} : \genstatecollocMatySlabArg{} \mapsto \begin{bmatrix}
 \sqrt{\quadWeightsLMSScalarArg{n}{1} } \stweightingMatOneArg{n,1} \residLMSArg{n,1}( \basisspaceArg{n} \genstateyDiscreteArgnt{1} + \stateInterceptArg{n},  \basisspaceArg{n} \genstateyDiscreteArgnt{0} + \stateInterceptArg{n},) \\
\vdots \\
 \sqrt{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}} } \stweightingMatOneArg{n,\nstepsArg{n}} \residLMSArg{n,\nstepsArg{n}}( \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}} + \stateInterceptArg{n}, \ldots,\basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n} - k^n(\nstepsArg{n}) } + \stateInterceptArg{n}) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{} \defeq \unroll (\genstateyDiscreteArg{0},\ldots,\genstateyDiscreteArg{\nstepsArg{n}} ).$
It is worth noting that, by design,
\begin{equation*}
\objectiveArgLMS{n} \bigg( \basisspaceArg{n} \genstateDiscreteArg{n,0} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateDiscreteArg{n,\nstepsArg{n}} + \stateInterceptArg{n} \ \bigg) 
\equiv
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard Gauss--Newton method. The algorithm consists of three fundamental steps: (1) compute the FOM O$\Delta$E residual given the current guess, (2) compute the time window residual-Jacobian and form the normal equations, and (3) solve the normal equations and update the state.

The practical implementation of the Gauss--Newton algorithm requires an efficient method for computing the time window residual-Jacobians. In particular, it is important to note that for linear multistep methods (and many other time marching schemes) the time window residual Jacobian will (almost) be a block-diagonal 
sparse matrix with the sparsity pattern,
\begin{equation*}
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n})  =
 \begin{bmatrix*}[l]
\matshapeb & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*}.
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
This sparsity pattern can be leveraged to assemble the time window residual-Jacobian from local residual Jacobians. 
Further, the sparsity pattern can be 
leveraged to efficiently compute the Jacobian matrix-matrix product in the normal equations.
It is also worth noting that the 
normal equations will also (almost) be block-diagonal; this is a fact that can be leveraged to speed up the linear solves at each Gauss--Newton iteration.

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
The principle cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations are several such examples. Further, the space--time formulation exposes an extra dimension for parallelization that can be exposed to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\text{tol}$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$k \leftarrow 0$ \Comment{Set counter}\\
$\genstatecollocMatSlabArg{n}_k \leftarrow \unroll(\genstateDiscreteArg{n,0},\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
\While{\text{converged} == \text{false}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
$\Delta \genstatecollocMatSlabArg{n} \solveAlgo [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
\If{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \text{tol}$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
$k\leftarrow k+1$
}
\end{algorithm}
